A full approximation storage multilevel method \\ for nonlinear variational inequalities

### 2023-06-28 paper/fascd.tex
fiddle with text ... this version for Patrick

Regarding the floating-point storage required, let us assuming all box constraints are nontrivial.  Then seven vectors must be allocated on each level: $\underline{\chi}^j$, $\overline{\chi}^j$, $\underline{\phi}^j$, $\overline{\phi}^j$, $w^j$, $\ell^j$, $y^j$.  Note $z^j$ can use the same storage as $y^j$.  On the finest level one must also store $\underline{\gamma}^J$ and $\overline{\gamma}^J$, while on the coarsest level there are only five vectors because $\underline{\phi}^0=\underline{\chi}^0$ and $\overline{\phi}^0=\overline{\chi}^0$.  Therefore, using $m_j=2^d m_{j-1}$, plus a geometric series argument, the total storage, excluding smoother and coarse-solver implementations, is
\begin{equation}
9 m_J + 7 m_{J-1} + \dots + 7 m_1 + 5 m_0 \le \left(9 + \frac{7}{2^d - 1}\right) m_J.
\end{equation}
This gives storage at most $16m_J$, $12m_J$, and $10m_J$ in dimensions $d=1,2,3$, respectively.  For comparison, a single-level method needs $4 m_J$ storage.\footnote{That is, for vectors $\underline{\gamma}^J,\overline{\gamma}^J,\ell^J,w^J$.}  If some of the obstacles are trivial, for instance in the unilateral case $\overline{\gamma}^J=+\infty$, storage is correspondingly reduced at all levels.

@@

In this section we consider several practical concerns when implementing and applying \pr{fascd-vcycle} Algorithm \ref{alg:fascd}.  We address convergence criteria, the F-cycle extension, and an active set Newton method used in Section \ref{sec:results} as a smoother and coarse solver.

Because the problem is inequality-constrained, repeated application of \pr{fascd-vcycle} to solve the finite-dimensional, box-constrained VI \eqref{eq:fe:vi} will not generally reduce the residual $f^J(w^J) - \ell^J$ to zero everywhere.  Indeed \eqref{eq:fe:vi} is equivalent to a \emph{mixed complementarity problem} (MCP) \cite{FacchineiPang2003}, a strong form amenable to constructing a convergence criteria for the discretization.  Let $x_p \in \mathcal{T}^J$ be a finest-level node, and $\psi_p(x)$ the corresponding hat function.  Let $\mathcal{X}^J=\big\{w\in\cV^J\,:\,w(x_p)=g_D^J(x_p) \text{ if } x_p\in \partial_D\Omega\big\}$, thus $\cK^J = \left\{\underline{\gamma}^J \le w \le \overline{\gamma}^J\right\} \cap \mathcal{X}^J$.  If $m_D$ is the number of nodes $x_p$ in the Dirichlet boundary, let $I_D$ be the $m_D\times m_J$ matrix which extracts Dirichlet nodal values; its transpose $I_D^\top$ injects Dirichlet boundary degrees of freedom into $(\mathcal{V}^J)'$.  Finally, write $a\perp b$ when $a,b \in \RR^n$ are \emph{complementary}, that is, when $a_i \ge 0$ and $b_i \ge 0$ and $a_i b_i = 0$ for $i=0,\dots,n-1$.  Then problem \eqref{eq:fe:vi}, which would be expressed VI$\left(f^J(\cdot)-\ell^J,\left[\,\underline{\gamma}^J,\overline{\gamma}^J\right] \cap \mathcal{X}^J\right)$ in the operations research literature \cite{FerrisPang1997}, is equivalent to the following MCP for the solution $u^J \in \mathcal{K}^J$, but with additional equality constraints:
\begin{align}
\underline{\gamma}^J \le u^J \le \overline{\gamma}^J \quad &\perp \quad f^J(u^J) - \ell^J - I_D^\top \lambda \label{eq:fe:mcp} \\
I_D u^J &= g_D^J\big|_{\partial_D\Omega} \notag
\end{align}
for some Lagrange multipliers $\lambda \in \RR^{m_D}$.

@@

We will instead measure convergence using a \emph{semi-smooth residual} based on the Fischer-Burmeister function \cite{BensonMunson2006,Ulbrich2011} on $a,b\in\RR$:
\begin{equation}
\phi(a,b) = a + b - \sqrt{a^2 + b^2}. \label{eq:phiFB}
\end{equation}
Observe that $\phi$ is semi-smooth, thus continuous, and that $\phi(a,b)=0$ if and only if $a,b$ are complementary.  For convenience denote $\underline{g}_p = w^J(x_p) - \underline{\gamma}^J(x_p)$ and $\overline{g}_p = \overline{\gamma}^J(x_p) - w^J(x_p)$, the nonnegative gaps at $x_p$.  Now define\footnote{This semismooth residual is straightforwardly modified for unconstrained degrees of freedom, that is, where $\overline{\gamma}^J(x_p) = +\infty$ or $\underline{\gamma}^J(x_p) = -\infty$ or both.}
\begin{equation}
\rSS(w^J)_p = \begin{cases}
\max\big\{\phi\big(\underline{g}_p, r(w^J)_p\big), \phi\left(\overline{g}_p, -r(w^J)_p\right)\big\}, & x_p \notin \partial_D\Omega, \\
w^J(x_p) - g_D^J(x_p), & x_p \in \partial_D\Omega.
\end{cases} \label{eq:rSS}
\end{equation}
Suppose an iteration starts with $w^{J,0} \in \mathcal{K}^J$, generating (fine-level) iterates $w^{J,k} \in \mathcal{K}^J$.  Let $\id{atol}\ge 0$, $\id{rtol}\ge 0$ be given tolerances.  Then the criterion used in Section \ref{sec:results} for computations is\footnote{A step tolerance $\|w^{J,k} - w^{J,k-1}\|_2 < \id{stol}$ or similar could also be used, but we did not find it necessary.}
\begin{equation}
\|\rSS(w^{J,k})\|_2 < \id{atol} \qquad \text{or} \qquad \frac{\|\rSS(w^{J,k})\|_2}{\|\rSS(w^{J,0})\|_2} < \id{rtol} \label{eq:stoppingcriterion}
\end{equation}

Straightforward iteration of V-cycle Algorithm \ref{alg:fascd}, starting with some finest-mesh initial iterate $w^{J,0}$ and repeated until \eqref{eq:stoppingcriterion} holds, is demonstrated as a solver in Section \ref{sec:results}.  However, application of an FASCD F-cycle, also called a \emph{full multigrid cycle}, is seen to be even better.  The well-known principle \cite[section 2.6]{Trottenbergetal2001} is to start on the coarsest mesh and solve the problem from any initial iterate.  Then an admissible initial iterate on the next-finer mesh is generated via prolongation and truncation, and so on.  During this ``ramp'' up one approximately solves on each level by using a fixed number (\id{rampv}) of V-cycles, each down to the coarsest mesh.  From the finest level one does full-depth V-cycles to convergence.

This F-cycle approach is in Figure \ref{fig:fcycle} and Algorithm \ref{alg:fascd-fcycle}.  Observe that injection of the original fine-level constraints $\underline{\gamma}^J, \overline{\gamma}^J$ is \emph{not} used in generating the LDCs for V-cycles; compare Section \ref{sec:cdmultilevel}.  However, injection is appropriate for the ramp in an F-cycle, if followed by truncation into $\mathcal{K}^j=\{v \in \mathcal{V}^j \,:\, \underline{\gamma}^j \le v \le \overline{\gamma}^j, \, v|_{\partial\Omega} = g_D\}$; see lines \ref{line:fcycleprolongtruncone} and \ref{line:fcycleprolongtrunctwo} in Algorithm \ref{alg:fascd-fcycle}.  Though not included in Algorithm \ref{alg:fascd-fcycle}, one may generate the coarsest-level initial iterate $w^0\in\mathcal{K}^0$ by injection and truncation down to coarsest level; compare line \ref{line:fcyclecoarsestinitial}.

\begin{figure}[h]
\begin{center}
\input{tikz/fcycle.tex}
\end{center}
\caption{Algorithm \ref{alg:fascd-fcycle} initially restricts the data $\ell^J,\underline{\gamma}^J,\overline{\gamma}^J$ down to level $j=0$ (gray edges).  Then it prolongs and truncates (doubled edges) to create an admissible initial iterate on the next-finer level.  Solution on each level is by V-cycles; \id{rampv} $=1$ here.  See Figure \ref{fig:fascdvcycle} for the other symbols.}
\label{fig:fcycle}
\end{figure}

@@

While a large space of smoothers could be tested,\footnote{See comments in Section \ref{sec:discussion}.} Section \ref{sec:results} only demonstrates one powerful, $O(m_j)$ work per iteration, smoother choice, namely a fixed number of iterations of the active-set Newton method from \cite{BensonMunson2006}, as implemented in PETSc \cite{Balayetal2023}, with a fixed number of preconditioned Krylov iterations to solve the reduced-space step equations.  We will denote this truncated active-set Newton method as \emph{RS Newton}.  The coarse solver is the same, but with LU decomposition solution of the step equations.  For the symmetric Laplacian in Example \ref{ex:results:classical} we used conjugate gradients (CG) with incomplete Cholesky (IC) preconditioning, but for other operators the generalized minimum residual (GMRES) method with incomplete LU (ILU) preconditioning is used.

>>>>

Regarding the floating-point storage required, let us assuming all box constraints are nontrivial.  Then seven vectors must be allocated on each level: $\underline{\chi}^j$, $\overline{\chi}^j$, $\underline{\phi}^j$, $\overline{\phi}^j$, $w^j$, $\ell^j$, $y^j$.  Note $z^j$ can use the same storage as $y^j$.  On the finest level one must also store $\underline{\gamma}^J$ and $\overline{\gamma}^J$, while on the coarsest level there are only five vectors because $\underline{\phi}^0=\underline{\chi}^0$ and $\overline{\phi}^0=\overline{\chi}^0$.  Therefore, using $m_j=2^d m_{j-1}$, plus a geometric series argument, the total storage, excluding smoother and coarse-solver implementations, is
\begin{equation}
9 m_J + 7 m_{J-1} + \dots + 7 m_1 + 5 m_0 \le \left(9 + \frac{7}{2^d - 1}\right) m_J.
\end{equation}
This gives storage at most $16m_J$, $12m_J$, and $10m_J$ in dimensions $d=1,2,3$, respectively.  For comparison, a single-level method needs $4 m_J$ storage (i.e.~$\underline{\gamma}^J,\overline{\gamma}^J,\ell^J,w^J$).  If some of the obstacles are trivial, for instance in the unilateral case $\overline{\gamma}^J=+\infty$, storage can be correspondingly reduced at all levels.

@@

In this section we consider the implementation of FASCD methods.  We address convergence criteria, the F-cycle extension, and an active set Newton method used as a smoother and coarse solver.

Because the problem is inequality-constrained, repeated application of Algorithm \ref{alg:fascd} to solve the finite-dimensional, box-constrained VI \eqref{eq:fe:vi} will not generally reduce the residual $f^J(w^J) - \ell^J$ to zero everywhere.  To construct an appropriate convergence criteria for the discretization, note \eqref{eq:fe:vi} is equivalent to a \emph{mixed complementarity problem} (MCP) \cite{FacchineiPang2003}, a strong form.  Let $x_p \in \mathcal{T}^J$ be a finest-level node, and $\psi_p(x)$ the corresponding hat function.  Let $\mathcal{X}^J=\big\{w\in\cV^J\,:\,w(x_p)=g_D^J(x_p) \text{ if } x_p\in \partial_D\Omega\big\}$, thus $\cK^J = \left\{\underline{\gamma}^J \le w \le \overline{\gamma}^J\right\} \cap \mathcal{X}^J$.  If $m_D$ is the number of nodes $x_p$ in the Dirichlet boundary, let $I_D$ be the $m_D\times m_J$ matrix which extracts Dirichlet nodal values; its transpose $I_D^\top$ injects Dirichlet boundary degrees of freedom into $(\mathcal{V}^J)'$.  Finally, write $a\perp b$ when $a,b \in \RR^n$ are \emph{complementary}, that is, when $a_i \ge 0$ and $b_i \ge 0$ and $a_i b_i = 0$ for $i=0,\dots,n-1$.  Then problem \eqref{eq:fe:vi}, which would be expressed VI$\left(f^J(\cdot)-\ell^J,\left[\,\underline{\gamma}^J,\overline{\gamma}^J\right] \cap \mathcal{X}^J\right)$ in the operations research literature \cite{FerrisPang1997}, is equivalent to the following MCP for the solution $u^J \in \mathcal{K}^J$, but with additional equality constraints:
\begin{align}
\underline{\gamma}^J \le u^J \le \overline{\gamma}^J \quad &\perp \quad f^J(u^J) - \ell^J - I_D^\top \lambda \label{eq:fe:mcp} \\
I_D u^J &= g_D^J\big|_{\partial_D\Omega} \notag
\end{align}
for some Lagrange multipliers $\lambda \in \RR^{m_D}$.

@@

We will instead measure convergence using a \emph{semi-smooth residual} based on the Fischer-Burmeister function \cite{BensonMunson2006,Ulbrich2011} on $a,b\in\RR$:
\begin{equation}
\phi(a,b) = a + b - \sqrt{a^2 + b^2}. \label{eq:phiFB}
\end{equation}
Observe that $\phi$ is semi-smooth, thus continuous, and that $\phi(a,b)=0$ if and only if $a,b$ are complementary.  For convenience denote $\underline{g}_p = w^J(x_p) - \underline{\gamma}^J(x_p)$ and $\overline{g}_p = \overline{\gamma}^J(x_p) - w^J(x_p)$, the nonnegative gaps at $x_p$.  Now define
\begin{equation}
\rSS(w^J)_p = \begin{cases}
\max\big\{\phi\big(\underline{g}_p, r(w^J)_p\big), \phi\left(\overline{g}_p, -r(w^J)_p\right)\big\}, & x_p \notin \partial_D\Omega, \\
w^J(x_p) - g_D^J(x_p), & x_p \in \partial_D\Omega.
\end{cases} \label{eq:rSS}
\end{equation}
Straightforward modifications apply for unconstrained degrees of freedom, where $\overline{\gamma}^J(x_p) = +\infty$ or $\underline{\gamma}^J(x_p) = -\infty$ or both.

Suppose an iteration starts with $w^{J,0} \in \mathcal{K}^J$, generating (fine-level) iterates $w^{J,k} \in \mathcal{K}^J$.  Let $\id{atol}\ge 0$, $\id{rtol}\ge 0$ be given tolerances.  We propose
\begin{equation}
\|\rSS(w^{J,k})\|_2 < \id{atol} \qquad \text{or} \qquad \frac{\|\rSS(w^{J,k})\|_2}{\|\rSS(w^{J,0})\|_2} < \id{rtol} \label{eq:stoppingcriterion}
\end{equation}
as the stopping criterion.  A step tolerance, $\|w^{J,k} - w^{J,k-1}\|_2 < \id{stol}$ or similar, could also be used, but for the results in Section \ref{sec:results} we did not find it necessary.

Iteration of V-cycle Algorithm \ref{alg:fascd}, starting with some finest-mesh initial iterate $w^{J,0}$ and repeated until \eqref{eq:stoppingcriterion} holds, will be shown (Section \ref{sec:results}) to be an effective solver.  However, an F-cycle (Figure \ref{fig:fcycle}), also called a \emph{full multigrid cycle}, is even better.  The well-known principle \cite[section 2.6]{Trottenbergetal2001} is to start on the coarsest mesh and solve the problem from any initial iterate.  Then an admissible initial iterate on the next-finer mesh is generated via prolongation and truncation, and so on.  During this ``ramp'' up one approximately solves on each level by using a fixed number of V-cycles, each down to the coarsest mesh.  From the finest level one does full-depth V-cycles to convergence.

\begin{figure}[h]
\begin{center}
\input{tikz/fcycle.tex}
\end{center}
\caption{F-cycle Algorithm \ref{alg:fascd-fcycle} initially restricts the data $\ell^J,\underline{\gamma}^J,\overline{\gamma}^J$ down to level $j=0$ (gray edges).  Then it prolongs and truncates (doubled edges) to create an admissible initial iterate on the next-finer level.  Solution on each level is by V-cycles; \id{rampv} $=1$ here.  See Figure \ref{fig:fascdvcycle} for the other symbols.}
\label{fig:fcycle}
\end{figure}

This F-cycle approach is in Algorithm \ref{alg:fascd-fcycle}.  Observe that while injection of the original fine-level constraints $\underline{\gamma}^J, \overline{\gamma}^J$ is \emph{not} used for V-cycles, wherein the LDCs use an existing admissible iterate (Section \ref{sec:cdmultilevel}), when combined with truncation into $\mathcal{K}^j=\{v \in \mathcal{V}^j \,:\, \underline{\gamma}^j \le v \le \overline{\gamma}^j, \, v|_{\partial\Omega} = g_D\}$ injection is appropriate for the ramp in an F-cycle; see lines \ref{line:fcycleprolongtruncone} and \ref{line:fcycleprolongtrunctwo}.  One may generate the coarsest-level initial iterate $w^0\in\mathcal{K}^0$ by injection and truncation down to coarsest level; compare line \ref{line:fcyclecoarsestinitial}.

@@

A large space of smoothers could be explored,\footnote{See comments in Section \ref{sec:discussion}.} but Section \ref{sec:results} only demonstrates one powerful, $O(m_j)$ work per iteration, smoother choice.  We apply a fixed number of iterations of the active-set (reduced-space) Newton method from \cite{BensonMunson2006}, the \texttt{-snes\_type vinewtonrsls} solver in PETSc \cite{Balayetal2023}, denoted as \emph{RS Newton}.  We use it with a fixed number of preconditioned Krylov iterations to solve the step equations.  For the symmetric Laplacian in Example \ref{ex:results:classical} we apply conjugate gradients (CG) with incomplete Cholesky (IC) preconditioning, but for other operators we apply the generalized minimum residual (GMRES) method with incomplete LU (ILU) preconditioning.  The coarse solver is also RS Newton, but iterated to convergence and with LU decomposition solution of the step equations.
